Are translation equivalents special? Evidence from simulations and empirical data from bilingual infants

### 2022-03-16 paper/knit to word/paper_TE_bilingual_vocabulary_model_220316.Rmd
fixed typos

header-includes:
  - \usepackage{amsmath}
  - \usepackage[labelformat=empty]{caption}
  - \usepackage{caption}
  - \renewcommand{\topfraction}{1}
  - \renewcommand{\bottomfraction}{1}
  - \renewcommand{\textfraction}{.1}
  - \renewcommand{\floatpagefraction}{1}
  - \setcounter{topnumber}{9}
  - \setcounter{bottomnumber}{9}
  - \setcounter{totalnumber}{20}
  - \setcounter{dbltopnumber}{9}
  - \DeclareUnicodeCharacter{2265}{\ensuremath{\ge}}
  - \DeclareUnicodeCharacter{2264}{\ensuremath{\le}}

@@

class             : "man"
output            : papaja::apa6_word # change to apa6_docx for word document
citation_package  : biblatex
nocite: '@*' # include all the references in the .bib file, because in-text citations were mostly typed manually to fit the apa7 style
---

@@

Bilingual children must learn words that take a different form in each of their languages, but share the same or highly similar meanings. For instance, to refer to the same crisp red-skinned fruit, an English–French bilingual child must use the word “apple” when speaking English, and the word _“pomme”_ when speaking French. These cross-language synonyms are known as translation equivalents (also called doublets; Umbel et al., 1992), and are observed amongst bilingual children’s first words (e.g., David & Wei, 2008; De Houwer et al., 2006; Pearson et al., 1995). Translation equivalents are thought to hold a special status in a bilingual’s developing lexicon due to the strong overlap in their semantics. For example, studies with bilingual toddlers show that the associative semantic properties of a word in one language facilitate the activation of its translation equivalent (e.g., Bilson et al., 2015; Floccia et al., 2020; Jardak & Byers-Heinlein, 2019). That is, upon hearing the English word “apple,” the corresponding French word _“pomme”_ is more easily activated in bilinguals’ minds. In vocabulary acquisition, bilingual children must learn a first label for a referent (a “singlet”; Umbel et al., 1992) before they can learn its translation equivalent (i.e., a second label for a referent, in a different language, learned following a singlet)[^1]. Is translation equivalent learning different from singlet learning? The current paper contrasts three competing accounts: 1) translation equivalents are harder to learn than singlets (Avoidance Account), 2) translation equivalents are easier to learn than singlets (Preference Account), and 3) translation equivalents and singlets are learned similarly (Neutral Account). To adjudicate between these accounts, we introduce the Bilingual Vocabulary Model, which provides a computational account of vocabulary learning, with parameters including bilinguals’ vocabulary in each language and their developmental level. In Study 1, we use the Bilingual Vocabulary Model to derive a set of predictions, which we then test against vocabulary data from 200 18- to 33-month-old bilingual children in Study 2.

@@

## Accounts of translation equivalent learning

### Avoidance Account: Translation equivalents are harder to learn than singlets
Early theories of bilingual development claimed that translation equivalents are conspicuously missing from bilingual children’s early vocabularies [e.g., @Imedadze_1967; @Swain_Wesche_1975; @Volterra_Taeschner_1978]. The phenomenon of missing translation equivalents led theorists to propose that young bilingual children do not differentiate their languages, and thus tend to learn only a single word for each referent. This avoidance of translation equivalents — particularly for nouns — was thought to be due to word learning biases such as mutual exclusivity, whereby children assume that a referent is only associated with one word at the basic level [@Markman_Wachtel_1988; @Markman_1992; @Markman_1994]. For example, when monolingual children see a familiar object (e.g., a cup) next to a novel object (e.g., a garlic press) and hear a novel word like “wug,” they assume that “wug” refers to the garlic press — the object unknown to them — rather than to the cup, the object for which they already know the word.

@@

### Preference Account: Translation equivalents are easier to learn than singlets
Contrary to the Avoidance Account, the Preference Account posits that translation equivalents are easier to learn than singlets. At a minimum, word learning requires encoding and representing the relevant sounds of a word, creating a mental representation of its referent, and linking the two. When a French–English bilingual child encounters the word _“pomme”_ after having learned “apple,” one part of that process has already occurred in that the referent is already represented; because part of the word learning task is already accomplished, translation equivalents might therefore be easier to learn than singlets (e.g., Montanari, 2010; Poulin-Dubois et al., 2013; 2018). Moreover, research suggests that bilingual lexicons are not tightly encapsulated by language, but instead include cross-language mental links between words that are semantically related [e.g., @Floccia_etal_2020; @Jardak_Byers-Heinlein_2019; @Singh_2014]. In this context, the strong semantic overlap makes translation equivalents special, and could facilitate their acquisition (e.g., Bilson et al., 2015; Floccia et al., 2020). The Preference Account predicts that translation equivalents are more easily learned than singlets.

@@

Other evidence suggesting translation equivalents might be easier to learn than singlets comes from a study that included vocabulary-checklist data from 254 monolingual and 181 bilingual children aged 6 months to 7.5 years (Bilson et al., 2015). The researchers used a network analysis approach to investigate how translation equivalents are learned, focusing on the semantic relationships between the words (e.g., words like “cat” and “dog” are strongly semantically related). Using a statistical model that allowed free semantic relations among vocabulary data from monolingual and bilingual children, the results suggested that words were learned faster when they were semantically connected to more known words in children’s lexicons. This effect applied not only to words within the same language, but also to words across languages including translation equivalents (e.g., English “dog” and French _“chien”_) and words that had other cross-language relations (e.g., “cat” and _“chien”_). The authors then simulated bilingual vocabulaires by modeling bilingual lexicons as combinations of two independent vocabulary-size-matched monolinguals. Comparison with actual bilingual children’s vocabulary data revealed that bilingual children acquired more translation equivalents than predicted by the simulation. The authors therefore concluded that bilingual children learn translation equivalents more easily than singlets. Note that, in their study, expected translation equivalent knowledge was simulated based on the number of lexical items that overlapped between two randomly chosen English monolinguals (e.g., whether both monolinguals knew the word “cat”). However, it is unclear whether this is an appropriate point of comparison for bilingual children as this approach may overlook variables that impact bilinguals’ vocabulary learning, including vocabulary size in each language and the developmental level of a child — a point that we will return to later in the introduction.

@@

### Neutral Account: Translation equivalents and singlets are learned similarly
The previous accounts rely on the idea that bilingual vocabulary development unfolds differently than monolingual development, as monolinguals encounter only singlets but bilinguals encounter both singlets and translation equivalents. There is an underlying assumption that translation equivalent learning is somehow special relative to singlet learning — the Avoidance Account proposes that translation equivalents are harder to learn than singlets, whereas the Preference Account proposes that translation equivalents are easier to learn than singlets. However, it is also possible that translation equivalents are neither harder nor easier for bilingual children to learn than singlets. We call this the Neutral Account.

@@

## Contributors to translation equivalent knowledge
The previous section discussed three theoretical accounts concerning the relative learnability of translation equivalents. However, to date, aspects of translation equivalent learning have mostly been examined in isolation, rather than integrated within the larger context of bilingual lexical development. In this section, we consider two proximal variables that we expect to predict the number of translation equivalents bilingual children know: vocabulary size in each language, and word learnability as a function of children’s developmental level.

### Vocabulary size in bilinguals’ two languages
Balance between the two vocabulary sizes is a function of the number of words bilingual children produce in each language, which tends to be tightly linked to their exposure to each language. In general, more language exposure leads to a larger vocabulary size (e.g., Barnes & Garcia, 2013; Boyce et al., 2013; Hurtado et al., 2014; Marchman et al., 2010; Pearson et al., 1997; Place & Hoff, 2011). Bilingual children usually know more words in the language to which they have greater exposure (i.e; dominant language) relative to the language to which they have less exposure (i.e., non-dominant language; Pearson et al., 1997; Place & Hoff, 2011). This is because the more often a bilingual hears a language, the more opportunities there will be for learning new words in that language.

@@

### Word learnability as a function of developmental level
An often overlooked factor that could contribute to bilingual children’s learning of translation equivalents is related to the changes in the learnability of different words over time based on children’s developmental level. Evidence from monolingual children shows that some types of words are characteristically learned before others. For example, across many languages including English, children show a noun bias in their early lexicons (Braginsky et al., 2019; Goodman et al., 2008), although for other languages such as Mandarin it appears that verbs and nouns are more equally acquired [@Tardif_1996]. Certain classes of words are rarely known at the onset of lexical development, such as prepositions and words for time [@Fenson_etal_2007]. This is thought to be due to the cognitive and linguistic machinery that must be in place in order for children to represent these concepts, a necessary prerequisite for learning certain word types (Bergelson, 2020; Braginsky et al., 2019). If this is the case, then children might be more likely to learn translation equivalents than singlets, simply because translation equivalents are more likely to be learnable at their stage of development. That is, many potential singlets could be “too hard” to learn at a particular age, while by definition potential translational equivalents refer to concepts that children are already able to lexicalize. For example, an 14-month-old who has learned the French words _“ballon”_ and _“chien”_ would be capable of learning their translation equivalents “ball” and “dog”, as well as easier potential singlets like “cup”, but not later-acquired potential singlets like “vitamins”, “tomorrow”, “who”, or “not”. Thus, a seeming overabundance of translation equivalents might be a product of developmental constraints on word learning, rather than due to semantic facilitation.

## The Bilingual Vocabulary Model
Taking into account the contributions of language exposure and developmental level to bilingual children’s vocabulary acquisition, we put forward the Bilingual Vocabulary Model. This model proposes that the number of translation equivalents that bilingual children produce is a function of vocabulary learning in each language, in the context of the number of potentially learnable words given the children’s developmental level. We formalize learning a translation equivalent pair as the joint probability of learning each of the words in the pair. This provides a straightforward empirical test of different theoretical accounts of translation equivalent learning, by asking whether or not the probability of knowing a word is independent of knowing its translation equivalent. The logic is similar to that of the familiar chi-squared test for independence, where the independence of two events from the same population is tested as the probability of their intersection computed by multiplying the probability of each individual event: P(A and B) = P(A) × P(B) if A and B are independent[^2]. In the next paragraphs, we define each of the model parameters in detail, and these are also summarized in Table 1.

@@

The model takes four main parameters as predictors: the number of words produced in the dominant language (DOM), the number of words produced in the non-dominant language (NONDOM), vocabulary size of potentially learnable words in each language (LEARNABLE), and a bias parameter (BIAS) which indicates whether the model is biased towards (BIAS > 1) or against (BIAS < 1) learning translation equivalents. The language in which a child knows more words is the dominant language, whereas the one in which a child knows fewer words is the non-dominant language. Next, we turn to the LEARNABLE parameter (i.e., the number of potentially learnable words). If DOM and NONDOM are measured with an instrument such as the MacArthur-Bates Communicative Development inventories (CDI; Fenson et al., 2007), one option would be to set LEARNABLE to be the total number of items on the CDI, which is typically 600–700 depending on the language of the adaptation. Yet, not all children will be capable of learning all of the CDI words. Typically, words that are more concrete occur more frequently in the input, and words that are associated with babies tend to be acquired earlier than other words (Braginsky et al., 2019). That is, very young children would not be expected to produce many of the advanced words on the CDI, such as “lawn mower,” “sidewalk,” or “vitamins”. Thus, a reasonable solution might be to determine how many CDI words are potentially learnable given the child’s developmental level, which could be approximated by their age. Since there are no CDI norms for French–English bilingual children, we will take the average English and French CDI norms to approximate how many words children might lexicalize in either language. For example, imagine that Jamie who is 18 months old produces 50 English words and 20 French words, thus a total of 70 words. Monolingual children his age with the very largest productive vocabularies (those at the 90th percentile averaging between English and French norms) produce a total of 240 words (retrieved from the Wordbank database version 0.3.1; Frank et al., 2016). Although there is likely considerable individual variability as to the cognitive capacity even amongst children of the same age, we argue that averaging between the English and French CDIs at the 90th percentile provides a reasonable — if imperfect — estimate of the number of learnable words (LEARNABLE) that a bilingual child of Jamie’s age could potentially acquire in each language. Thus, we might expect that Jamie could potentially have learned up to 240 words in English and 240 words in French, although he has thus far only learned 50 in English and 20 in French. Note that we model learnability in terms of the number of words expected to be learnable at a particular age, rather than modeling the learnability of individual words at a particular age — a simplifying choice that we will return to in the discussion section.

@@

table1 <- data.frame(
  Variable = c("LEARNABLE", "DOM", "NONDOM", "BIAS", "BALANCE", "WORD", "TE", "CONCEPT", "DOM-SINGLET", "NONDOM-SINGLET"),
  Definition = c("Number of words expected to be learnable in each language, given the child’s developmental level",
                 "Words produced in the dominant language",
                 "Words produced in the non-dominant language",
                 "Bias parameter",
                 "Balance (relative proportion of words produced in the non-dominant language to the total words produced in both languages)",
                 "Word vocabulary (or total vocabulary size)",
                 "Translation equivalents produced",
                 "Concept vocabulary (or total conceptual vocabulary size)",
                 "Singlets in dominant language",
                 "Singlets in non-dominant language"),
  Constraints = c("Varies by age. No greater than the number of words on CDI.",
                  "DOM \u2265 NONDOM (children always produce more words in dominant than non-dominant language); DOM \u2264 LEARNABLE",
                  "NONDOM \u2264 DOM (children always produce fewer words in non-dominant than dominant language); NONDOM \u2264 LEARNABLE",
                  "BIAS < 1 implies the Avoidance Account; BIAS > 1 implies the Preference Account; BIAS = 1 implies the Neutral Account",
                  "0 \u2264 BALANCE \u2264 .50 (greater values indicate children producing a more similar number of words in their two languages)",
                  "W \u2264 2×LEARNABLE (maximum word vocabulary is knowing each word in both languages)",
                  "", "", "", ""),
  "Relationship to other parameters" = c("Maximum number that could be learned in DOM or NONDOM",
                                         "DOM = (1-BALANCE)×WORD; DOM = WORD - NONDOM",
                                         "NONDOM = BALANCE×WORD; NONDOM = WORD - DOM",
                                         "",
                                         "BALANCE = NONDOM/WORD; BALANCE = NONDOM/(DOM+NONDOM)",
                                         "WORD = DOM+NONDOM; WORD = DOM/(1-BALANCE); WORD = NONDOM/(BALANCE)",
                                         "TE = BIAS × DOM×NONDOM/LEARNABLE",
                                         "CONCEPT = WORD - TE; CONCEPT = TE + DOM-SINGLET + NONDOM-SINGLET",
                                         "DOM-SINGLET = DOM - TE",
                                         "NONDOM_SINGLET = NONCOM - TE")
)

@@

## Current studies
The current studies aimed to better understand the nature of translation equivalent learning in bilingual children. Study 1 simulated the expected patterns of translation equivalent learning under the Bilingual Vocabulary Model proposed in the introduction, with reference to the proportion of words learned in the dominant and non-dominant language and the number of words that are learnable at various developmental levels. We also compared predicted learning outcomes for when translation equivalent learning is harder than, easier than, or similar to singlet learning.

@@

# Study 1: Simulations
Study 1 provides a computational implementation of the Bilingual Vocabulary Model outlined in the introduction, which we use to simulate different scenarios to examine the effect of vocabulary sizes and developmental variables on translation equivalent learning. Simulations were conducted by holding constant or varying the main parameters across a set of possible values, and computing expected translation equivalents and derived parameters from these values. The number of children in each simulation was fully determined by the number of levels varied for each parameter: each combination of parameters was included, except those that violated constraints in the model (e.g., NONDOM could not be larger than DOM).

For each simulation, we generated predictions for expected patterns in bilingual children’s vocabulaires. Several previous studies in the literature have examined relations between translation equivalent knowledge and our main parameters including age or developmental level (which determines LEARNABLE; David & Wei, 2008; Legacy et al., 2016), DOM (Legacy et al., 2016; Poulin-Dubois et al., 2013), and NON-DOM (Legacy et al., 2016). However, others have examined derived parameters such as BALANCE (David & Wei, 2008; Legacy et al., 2016; Montanari, 2010; Pearson et al., 1995; 1997), WORD (Legacy et al., 2016; Montanari, 2010), CONCEPT (Pearson et al., 1993), DOM-SINGLET and NONDOM-SINGLET (De Houwer et al., 2006; Pearson et al., 1993; 1995). To facilitate comparison with the extant literature where main parameters are not always reported, we generated predictions expressed both in terms of main parameters and derived parameters. This provides alternate ways of testing the model, depending on which parameters are available in the dataset. Note that predictions from the main parameters and those from derived parameters do not provide independent tests of our model for the same dataset, as they can be algebraically calculated from one another. As an analogy, a child’s age is not independent from their birth date and today’s date, yet we might make reference to their age in some situations and their birth date in others. Similarly, elucidating relations among both main and derived parameters provides a more complete description of expected patterns in bilingual vocabulary development.

@@

## 1.1 Simulation 1: Children of the same developmental level with different word vocabularies and balances of vocabulary
In Simulation 1, we examined patterns of translation equivalent learning where simulated children are at the same developmental level and thus have the same number of potentially LEARNABLE words, but their vocabulary size in each language (DOM and NONDOM) varied. For readers less familiar with this type of approach, we provide expanded examples of three simulated infants in the Appendix. For convenience, we set LEARNABLE constant at 600, which roughly corresponds to what is expected for an English–French 27-month-old bilingual child. For example, a 27-month-old English-learner at the 90th percentile of vocabulary produces around 633 words and a 27-month-old French-learner produces around 580 words as retrieved from the Wordbank database, giving us an average of 606 words across the two languages (Frank et al., 2016). BIAS was set constant at 1, meaning that in these simulations translation equivalents and singlets are equally easy to learn. Data from a total of 216 simulated children were generated (see Table 2 for a summary of the parameter values used in this simulation). Based on these values, we derived simulated children’s word vocabulary (WORD, calculated as DOM+NONDOM) and their vocabulary balance (BALANCE, calculated as NONDOM/(DOM+NONDOM)).

@@

# Study 2: Empirical data
In Study 1, we used simulations based on the Bilingual Vocabulary Model to generate several predictions about the relationship between translation equivalent knowledge and other vocabulary variables. In Study 2, we tested these predictions using archival vocabulary data from 200 French–English bilingual children aged 18 to 33 months.

>>>>

header-includes:
  - \usepackage{amsmath}
  - \usepackage[labelformat=empty]{caption}
  - \usepackage{caption}
  - \renewcommand{\topfraction}{1}
  - \renewcommand{\bottomfraction}{1}
  - \renewcommand{\textfraction}{.1}
  - \renewcommand{\floatpagefraction}{1}
  - \setcounter{topnumber}{9}
  - \setcounter{bottomnumber}{9}
  - \setcounter{totalnumber}{20}
  - \setcounter{dbltopnumber}{9}
  - \DeclareUnicodeCharacter{2265}{\ensuremath{\ge}}
  - \DeclareUnicodeCharacter{2264}{\ensuremath{\le}}
  - \DeclareUnicodeCharacter{2212}{\ensuremath{−}}

@@

class             : "man"
output            : papaja::apa6_docx # change to apa6_docx for word document
citation_package  : biblatex
nocite: '@*' # include all the references in the .bib file, because in-text citations were mostly typed manually to fit the apa7 style
---

@@

# 1 Introduction
Bilingual children must learn words that take a different form in each of their languages, but share the same or highly similar meanings. For instance, to refer to the same crisp red-skinned fruit, an English–French bilingual child must use the word “apple” when speaking English, and the word _“pomme”_ when speaking French. These cross-language synonyms are known as translation equivalents (also called doublets; Umbel et al., 1992), and are observed amongst bilingual children’s first words (e.g., David & Wei, 2008; De Houwer et al., 2006; Pearson et al., 1995). Translation equivalents are thought to hold a special status in a bilingual’s developing lexicon due to the strong overlap in their semantics. For example, studies with bilingual toddlers show that the associative semantic properties of a word in one language facilitate the activation of its translation equivalent (e.g., Bilson et al., 2015; Floccia et al., 2020; Jardak & Byers-Heinlein, 2019). That is, upon hearing the English word “apple,” the corresponding French word _“pomme”_ is more easily activated in bilinguals’ minds. In vocabulary acquisition, bilingual children must learn a first label for a referent (a “singlet”; Umbel et al., 1992) before they can learn its translation equivalent (i.e., a second label for a referent, in a different language, learned following a singlet)[^1]. Is translation equivalent learning different from singlet learning? The current paper contrasts three competing accounts: 1) translation equivalents are harder to learn than singlets (Avoidance Account), 2) translation equivalents are easier to learn than singlets (Preference Account), and 3) translation equivalents and singlets are learned similarly (Neutral Account). To adjudicate between these accounts, we introduce the Bilingual Vocabulary Model, which provides a computational account of vocabulary learning, with parameters including bilinguals’ vocabulary in each language and their developmental level. In Study 1, we use the Bilingual Vocabulary Model to derive a set of predictions, which we then test against vocabulary data from 200 18- to 33-month-old bilingual children in Study 2.

@@

## 1.1 Accounts of translation equivalent learning

### 1.1.1 Avoidance Account: Translation equivalents are harder to learn than singlets
Early theories of bilingual development claimed that translation equivalents are conspicuously missing from bilingual children’s early vocabularies [e.g., @Imedadze_1967; @Swain_Wesche_1975; @Volterra_Taeschner_1978]. The phenomenon of missing translation equivalents led theorists to propose that young bilingual children do not differentiate their languages, and thus tend to learn only a single word for each referent. This avoidance of translation equivalents — particularly for nouns — was thought to be due to word learning biases such as mutual exclusivity, whereby children assume that a referent is only associated with one word at the basic level [@Markman_Wachtel_1988; @Markman_1992; @Markman_1994]. For example, when monolingual children see a familiar object (e.g., a cup) next to a novel object (e.g., a garlic press) and hear a novel word like “wug,” they assume that “wug” refers to the garlic press — the object unknown to them — rather than to the cup, the object for which they already know the word.

@@

### 1.1.2 Preference Account: Translation equivalents are easier to learn than singlets
Contrary to the Avoidance Account, the Preference Account posits that translation equivalents are easier to learn than singlets. At a minimum, word learning requires encoding and representing the relevant sounds of a word, creating a mental representation of its referent, and linking the two. When a French–English bilingual child encounters the word _“pomme”_ after having learned “apple,” one part of that process has already occurred in that the referent is already represented; because part of the word learning task is already accomplished, translation equivalents might therefore be easier to learn than singlets (e.g., Montanari, 2010; Poulin-Dubois et al., 2013; 2018). Moreover, research suggests that bilingual lexicons are not tightly encapsulated by language, but instead include cross-language mental links between words that are semantically related [e.g., @Floccia_etal_2020; @Jardak_Byers-Heinlein_2019; @Singh_2014]. In this context, the strong semantic overlap makes translation equivalents special, and could facilitate their acquisition (e.g., Bilson et al., 2015; Floccia et al., 2020). The Preference Account predicts that translation equivalents are more easily learned than singlets.

@@

Other evidence suggesting translation equivalents might be easier to learn than singlets comes from a study that included vocabulary-checklist data from 254 monolingual and 181 bilingual children aged 6 months to 7.5 years (Bilson et al., 2015). The researchers used a network analysis approach to investigate how translation equivalents are learned, focusing on the semantic relationships between the words (e.g., words like “cat” and “dog” are strongly semantically related). Using a statistical model that allowed free semantic relations among vocabulary data from monolingual and bilingual children, the results suggested that words were learned faster when they were semantically connected to more known words in children’s lexicons. This effect applied not only to words within the same language, but also to words across languages including translation equivalents (e.g., English “dog” and French _“chien”_) and words that had other cross-language relations (e.g., “cat” and _“chien”_). The authors then simulated bilingual vocabularies by modeling bilingual lexicons as combinations of two independent vocabulary-size-matched monolinguals. Comparison with actual bilingual children’s vocabulary data revealed that bilingual children acquired more translation equivalents than predicted by the simulation. The authors therefore concluded that bilingual children learn translation equivalents more easily than singlets. Note that, in their study, expected translation equivalent knowledge was simulated based on the number of lexical items that overlapped between two randomly chosen English monolinguals (e.g., whether both monolinguals knew the word “cat”). However, it is unclear whether this is an appropriate point of comparison for bilingual children as this approach may overlook variables that impact bilinguals’ vocabulary learning, including vocabulary size in each language and the developmental level of a child — a point that we will return to later in the introduction.

@@

### 1.1.3 Neutral Account: Translation equivalents and singlets are learned similarly
The previous accounts rely on the idea that bilingual vocabulary development unfolds differently than monolingual development, as monolinguals encounter only singlets but bilinguals encounter both singlets and translation equivalents. There is an underlying assumption that translation equivalent learning is somehow special relative to singlet learning — the Avoidance Account proposes that translation equivalents are harder to learn than singlets, whereas the Preference Account proposes that translation equivalents are easier to learn than singlets. However, it is also possible that translation equivalents are neither harder nor easier for bilingual children to learn than singlets. We call this the Neutral Account.

@@

## 1.2 Contributors to translation equivalent knowledge
The previous section discussed three theoretical accounts concerning the relative learnability of translation equivalents. However, to date, aspects of translation equivalent learning have mostly been examined in isolation, rather than integrated within the larger context of bilingual lexical development. In this section, we consider two proximal variables that we expect to predict the number of translation equivalents bilingual children know: vocabulary size in each language, and word learnability as a function of children’s developmental level.

### 1.1.1 Vocabulary size in bilinguals’ two languages
Balance between the two vocabulary sizes is a function of the number of words bilingual children produce in each language, which tends to be tightly linked to their exposure to each language. In general, more language exposure leads to a larger vocabulary size (e.g., Barnes & Garcia, 2013; Boyce et al., 2013; Hurtado et al., 2014; Marchman et al., 2010; Pearson et al., 1997; Place & Hoff, 2011). Bilingual children usually know more words in the language to which they have greater exposure (i.e; dominant language) relative to the language to which they have less exposure (i.e., non-dominant language; Pearson et al., 1997; Place & Hoff, 2011). This is because the more often a bilingual hears a language, the more opportunities there will be for learning new words in that language.

@@

### 1.1.2 Word learnability as a function of developmental level
An often overlooked factor that could contribute to bilingual children’s learning of translation equivalents is related to the changes in the learnability of different words over time based on children’s developmental level. Evidence from monolingual children shows that some types of words are characteristically learned before others. For example, across many languages including English, children show a noun bias in their early lexicons (Braginsky et al., 2019; Goodman et al., 2008), although for other languages such as Mandarin it appears that verbs and nouns are more equally acquired [@Tardif_1996]. Certain classes of words are rarely known at the onset of lexical development, such as prepositions and words for time [@Fenson_etal_2007]. This is thought to be due to the cognitive and linguistic machinery that must be in place in order for children to represent these concepts, a necessary prerequisite for learning certain word types (Bergelson, 2020; Braginsky et al., 2019). If this is the case, then children might be more likely to learn translation equivalents than singlets, simply because translation equivalents are more likely to be learnable at their stage of development. That is, many potential singlets could be “too hard” to learn at a particular age, while by definition potential translational equivalents refer to concepts that children are already able to lexicalize. For example, an 14-month-old who has learned the French words _“ballon”_ and _“chien”_ would be capable of learning their translation equivalents “ball” and “dog”, as well as easier potential singlets like “cup”, but not later-acquired potential singlets like “vitamins”, “tomorrow”, “who”, or “not”. Thus, a seeming overabundance of translation equivalents might be a product of developmental constraints on word learning, rather than due to semantic facilitation.

## 1.3 The Bilingual Vocabulary Model
Taking into account the contributions of language exposure and developmental level to bilingual children’s vocabulary acquisition, we put forward the Bilingual Vocabulary Model. This model proposes that the number of translation equivalents that bilingual children produce is a function of vocabulary learning in each language, in the context of the number of potentially learnable words given the children’s developmental level. We formalize learning a translation equivalent pair as the joint probability of learning each of the words in the pair. This provides a straightforward empirical test of different theoretical accounts of translation equivalent learning, by asking whether or not the probability of knowing a word is independent of knowing its translation equivalent. The logic is similar to that of the familiar chi-squared test for independence, where the independence of two events from the same population is tested as the probability of their intersection computed by multiplying the probability of each individual event: P(A and B) = P(A) × P(B) if A and B are independent[^2]. In the next paragraphs, we define each of the model parameters in detail, and these are also summarized in Table 1.

@@

The model takes four main parameters as predictors: the number of words produced in the dominant language (DOM), the number of words produced in the non-dominant language (NONDOM), vocabulary size of potentially learnable words in each language (LEARNABLE), and a bias parameter (BIAS) which indicates whether the model is biased towards (BIAS > 1) or against (BIAS < 1) learning translation equivalents. The language in which a child knows more words is the dominant language, whereas the one in which a child knows fewer words is the non-dominant language. Next, we turn to the LEARNABLE parameter (i.e., the number of potentially learnable words). If DOM and NONDOM are measured with an instrument such as the MacArthur-Bates Communicative Development Inventories (CDI; Fenson et al., 2007), one option would be to set LEARNABLE to be the total number of items on the CDI, which is typically 600–700 depending on the language of the adaptation. Yet, not all children will be capable of learning all of the CDI words. Typically, words that are more concrete occur more frequently in the input, and words that are associated with babies tend to be acquired earlier than other words (Braginsky et al., 2019). That is, very young children would not be expected to produce many of the advanced words on the CDI, such as “lawn mower,” “sidewalk,” or “vitamins”. Thus, a reasonable solution might be to determine how many CDI words are potentially learnable given the child’s developmental level, which could be approximated by their age. Since there are no CDI norms for French–English bilingual children, we will take the average English and French CDI norms to approximate how many words children might lexicalize in either language. For example, imagine that Jamie who is 18 months old produces 50 English words and 20 French words, thus a total of 70 words. Monolingual children his age with the very largest productive vocabularies (those at the 90th percentile averaging between English and French norms) produce a total of 240 words (retrieved from the Wordbank database version 0.3.1; Frank et al., 2016). Although there is likely considerable individual variability as to the cognitive capacity even amongst children of the same age, we argue that averaging between the English and French CDIs at the 90th percentile provides a reasonable — if imperfect — estimate of the number of learnable words (LEARNABLE) that a bilingual child of Jamie’s age could potentially acquire in each language. Thus, we might expect that Jamie could potentially have learned up to 240 words in English and 240 words in French, although he has thus far only learned 50 in English and 20 in French. Note that we model learnability in terms of the number of words expected to be learnable at a particular age, rather than modeling the learnability of individual words at a particular age — a simplifying choice that we will return to in the discussion section.

@@

table1 <- data.frame(
  Variable = c("LEARNABLE", "DOM", "NONDOM", "BIAS", "BALANCE", "WORD", "TE", "CONCEPT", "DOM-SINGLET", "NONDOM-SINGLET"),
  Definition = c("Number of words expected to be learnable in each language, given the child’s developmental level",
                 "Words produced in the dominant language",
                 "Words produced in the non-dominant language",
                 "Bias parameter",
                 "Balance (relative proportion of words produced in the non-dominant language to the total words produced in both languages)",
                 "Word vocabulary (or total vocabulary size)",
                 "Translation equivalents produced",
                 "Concept vocabulary (or total conceptual vocabulary size)",
                 "Singlets in dominant language",
                 "Singlets in non-dominant language"),
  Constraints = c("Varies by age. No greater than the number of words on CDI.",
                  "DOM \u2265 NONDOM (children always produce more words in dominant than non-dominant language); DOM \u2264 LEARNABLE",
                  "NONDOM \u2264 DOM (children always produce fewer words in non-dominant than dominant language); NONDOM \u2264 LEARNABLE",
                  "BIAS < 1 implies the Avoidance Account; BIAS > 1 implies the Preference Account; BIAS = 1 implies the Neutral Account",
                  "0 \u2264 BALANCE \u2264 .50 (greater values indicate children producing a more similar number of words in their two languages)",
                  "W \u2264 2×LEARNABLE (maximum word vocabulary is knowing each word in both languages)",
                  "", "", "", ""),
  "Relationship to other parameters" = c("Maximum number that could be learned in DOM or NONDOM",
                                         "DOM = (1 – BALANCE)×WORD; DOM = WORD – NONDOM",
                                         "NONDOM = BALANCE×WORD; NONDOM = WORD – DOM",
                                         "",
                                         "BALANCE = NONDOM/WORD; BALANCE = NONDOM/(DOM+NONDOM)",
                                         "WORD = DOM+NONDOM; WORD = DOM/(1 – BALANCE); WORD = NONDOM/(BALANCE)",
                                         "TE = BIAS × DOM×NONDOM/LEARNABLE",
                                         "CONCEPT = WORD – TE; CONCEPT = TE + DOM-SINGLET + NONDOM-SINGLET",
                                         "DOM-SINGLET = DOM – TE",
                                         "NONDOM-SINGLET = NONCOM – TE")
)

@@

## 1.4 Current studies
The current studies aimed to better understand the nature of translation equivalent learning in bilingual children. Study 1 simulated the expected patterns of translation equivalent learning under the Bilingual Vocabulary Model proposed in the introduction, with reference to the proportion of words learned in the dominant and non-dominant language and the number of words that are learnable at various developmental levels. We also compared predicted learning outcomes for when translation equivalent learning is harder than, easier than, or similar to singlet learning.

@@

# 2 Study 1: Simulations
Study 1 provides a computational implementation of the Bilingual Vocabulary Model outlined in the introduction, which we use to simulate different scenarios to examine the effect of vocabulary sizes and developmental variables on translation equivalent learning. Simulations were conducted by holding constant or varying the main parameters across a set of possible values, and computing expected translation equivalents and derived parameters from these values. The number of children in each simulation was fully determined by the number of levels varied for each parameter: each combination of parameters was included, except those that violated constraints in the model (e.g., NONDOM could not be larger than DOM).

For each simulation, we generated predictions for expected patterns in bilingual children’s vocabularies. Several previous studies in the literature have examined relations between translation equivalent knowledge and our main parameters including age or developmental level (which determines LEARNABLE; David & Wei, 2008; Legacy et al., 2016), DOM (Legacy et al., 2016; Poulin-Dubois et al., 2013), and NON-DOM (Legacy et al., 2016). However, others have examined derived parameters such as BALANCE (David & Wei, 2008; Legacy et al., 2016; Montanari, 2010; Pearson et al., 1995; 1997), WORD (Legacy et al., 2016; Montanari, 2010), CONCEPT (Pearson et al., 1993), DOM-SINGLET and NONDOM-SINGLET (De Houwer et al., 2006; Pearson et al., 1993; 1995). To facilitate comparison with the extant literature where main parameters are not always reported, we generated predictions expressed both in terms of main parameters and derived parameters. This provides alternate ways of testing the model, depending on which parameters are available in the dataset. Note that predictions from the main parameters and those from derived parameters do not provide independent tests of our model for the same dataset, as they can be algebraically calculated from one another. As an analogy, a child’s age is not independent from their birth date and today’s date, yet we might make reference to their age in some situations and their birth date in others. Similarly, elucidating relations among both main and derived parameters provides a more complete description of expected patterns in bilingual vocabulary development.

@@

## 2.1 Simulation 1: Children of the same developmental level with different word vocabularies and balances of vocabulary
In Simulation 1, we examined patterns of translation equivalent learning where simulated children are at the same developmental level and thus have the same number of potentially LEARNABLE words, but their vocabulary size in each language (DOM and NONDOM) varied. For readers less familiar with this type of approach, we provide expanded examples of three simulated infants in the Appendix. For convenience, we set LEARNABLE constant at 600, which roughly corresponds to what is expected for an English–French 27-month-old bilingual child. For example, a 27-month-old English-learner at the 90th percentile of vocabulary produces around 633 words and a 27-month-old French-learner produces around 580 words as retrieved from the Wordbank database, giving us an average of 606 words across the two languages (Frank et al., 2016). BIAS was set constant at 1, meaning that in these simulations translation equivalents and singlets are equally easy to learn. Data from a total of 216 simulated children were generated (see Table 2 for a summary of the parameter values used in this simulation). Based on these values, we derived simulated children’s word vocabulary (WORD, calculated as DOM+NONDOM) and their vocabulary balance (BALANCE, calculated as NONDOM/(DOM+NONDOM)).

@@

# 3 Study 2: Empirical data
In Study 1, we used simulations based on the Bilingual Vocabulary Model to generate several predictions about the relationship between translation equivalent knowledge and other vocabulary variables. In Study 2, we tested these predictions using archival vocabulary data from 200 French–English bilingual children aged 18 to 33 months.
